---
title: "Hospital wastewater virome"
author: "Eevi Sipponen"
date: 'Aug 13 2026'
output: html_document
---

Script for producing Figure 1 and Suppelementary Figure 1.

```{r define-colors-to-use}
site_cat_colors <- c("WWTP" = "#2E86C1",
                     "UH" = "#E67E22",
                     "SA" = "#42a16a")


other_colors <- c("#CC79A7", "#D55E00", "#56B4E9", "#332288", "#44AA99", 
                  "#AA4499", "#117733", "#88CCEE", "#F0E442", "#000000")
```

## Load libraries

```{r message=FALSE, warning=FALSE}
library(data.table)
library(openxlsx)
library(ggplot2)
library(ggtree)
library(cowplot)
library(ade4)
library(dplyr)
library(vegan)
library(tidyr)
library(DescTools)
library(stringr)
```

## Define path to use

User defined path containing required files to produce figures
- EsViritu_abundance_metadata.tsv

Profuced figures are set to be saved to the same location.

```{r}
path = "path/to/files"
```


## Load data

```{r}
data_file <- file.path(path, "EsViritu_abundance_metadata.tsv")

esviritu_passive_filt <- fread(data_file) %>%
  mutate(Site_cat = case_when(str_detect(sample_ID, "WWTP") ~ "WWTP",
                            str_detect(sample_ID, "UH") ~ "UH",
                            str_detect(sample_ID, "SA") ~ "SA",
                            TRUE ~ NA_character_),
         City = case_when(str_detect(sample_ID, "TRE") ~ "TRE",
                            str_detect(sample_ID, "KUO") ~ "KUO",
                            TRUE ~ NA_character_)) %>%
  filter(host_species != "Contamination")

```

## Supplementary figure S1

### Mapped reads and viral read proportions

Viral reads and viral read proportion of all reads
```{r}
summarize_viral_reads <- function(data){
  data %>%
    group_by(sample_ID) %>%
    summarize(total_mapped=sum(read_count),
              total_reads=first(filtered_reads_in_sample),
              viral_reads_percentage=(total_mapped/total_reads)*100,
              .groups="drop")
}
```

```{r}
#plot viral read proportion of all samples
plot_S1A <- function(data, colors){
  ggplot(data=data, aes(x="",y=viral_reads_percentage)) +
    geom_boxplot(fill=colors) +
   # scale_fill_manual(values=colors) +
    stat_summary(fun="median", geom="text", aes(label=signif(after_stat(y),2), vjust=-0.5)) +
    theme_bw()+
    labs(y = "Percentage",
         x="") +
    theme(axis.text.y = element_text(size=10),
          title = element_text(size=10),
          legend.position = "none") + 
    scale_y_continuous(labels = function(x) paste0(x, "%"))

}


#plot number of viral reads for all samples
plot_S1B <- function(data, colors){
  ggplot(data=data, aes(x="",y=total_mapped)) +
    geom_boxplot(fill=colors) +
    #scale_fill_manual(values=colors) +
    stat_summary(fun="median", geom="text", aes(label=round(after_stat(y),2), vjust=-0.5)) +
    theme_bw()+
    labs(y = "Reads aligned to viral genomes (N)",
         x="") +
    scale_y_log10() +
    theme(axis.text.x = element_text(angle=45),
          axis.text.y = element_text(size=10),
          title = element_text(size=10),
          legend.position = "none")
}
```

```{r}
v_reads_all <- summarize_viral_reads(esviritu_passive_filt)

Fig_S1A <- plot_S1A(v_reads_all, other_colors[1])

Fig_S1B <- plot_S1B(v_reads_all,other_colors[1])

plot_grid(Fig_S1A, Fig_S1B, ncol=2)

ggsave(plot_grid(Fig_S1A, Fig_S1B, ncol=2), 
       file = file.path(data_dir, "Fig_S1AB.jpg"), 
       width = 6, height = 4, units = "in", dpi=300)
```

### Assembly coverage boxplots

```{r}
esviritu_passive_meta_coverage <- esviritu_passive_filt  %>%
  mutate(coverage = covered_bases/Asm_length) %>%
  mutate(cov_category = case_when(
    coverage >= 0.9 ~ "Over 90%",
    coverage < 0.9 & coverage >= 0.5 ~ "50% to 90%",
    coverage < 0.5 & coverage >= 0.1 ~ "10% to 50%",
    TRUE ~ "< 10%"
  ))
```

```{r}
median_category <- esviritu_passive_meta_coverage %>%
  group_by(sample_ID, cov_category) %>%
  summarize(seqs = n_distinct(Assembly)) %>%
  ungroup() %>%
  group_by(cov_category) %>%
  summarize(median=median(seqs))


violin_data <- esviritu_passive_meta_coverage %>%
  group_by(sample_ID, cov_category) %>%
  summarize(seqs = n_distinct(Assembly)) %>%
  group_by(cov_category) %>%
  mutate(cov_category = factor(cov_category, levels = c("< 10%",
                                                        "10% to 50%",
                                                        "50% to 90%",
                                                        "Over 90%")))

Fig_S1C <- ggplot(data=violin_data, aes(x=seqs, y=cov_category, fill = cov_category, alpha = 0.7)) +
  geom_violin(alpha=0.6,color = "grey20", draw_quantiles = c(0.5)) +
  scale_fill_manual(values = other_colors) +
  scale_color_manual(values = other_colors) +
  stat_summary(fun = median, geom = "text", aes(label = round(..x.., 1)), angle=90,, vjust=1.2)+
  xlim(c(0,170)) +
  theme_bw() +
  labs(x = "Reference assemblies\nwith coverage level per sample (N)",
       y = "Assembly coverage") +
  theme(legend.position = "Off",
        axis.text.x = element_text(size=12),
        strip.text = element_text(size=12),
        text = element_text(size=14))


Fig_S1C

ggsave(Fig_S1C, 
       file = file.path(data_dir, "Fig_S1C"), 
       width = 7.2, height = 4, units = "in", dpi=300)
```
### Taxa accumulation plot

```{r}
RPKMF_rank <- function(data, tax_level){
  RPKMF_df <- data %>%
    group_by(sample_ID, !!sym(tax_level)) %>%
    #sum tax level RPKMF values at sample level
    summarize(RPKMF = sum(RPKMF_norm)) %>%
    ungroup()
  
  #transform long df to wider format, replace missing RPKMF values with 0
  RPKMF_df_long <- pivot_wider(data=RPKMF_df,
                    names_from = sample_ID,
                    values_from = RPKMF,
                    values_fill = list(RPKMF=0))
  
  
  return(RPKMF_df_long)
}

species_accumulation <- function(data, taxa_level){
  data <- as.data.frame(data[,c(-1)])
  rownames(data) <- data[[taxa_level]]
  
  accumulation <- specaccum(t(data), method="random", permutations=1000)
  boxplot_data <- as.data.frame(t(accumulation$perm))
  
  return(boxplot_data)
  
}

plot_S1D <- function(species_data, subspecies_data, genera_data){

    boxplot_data_s <- species_accumulation(species_data, "species")
    boxplot_data_ss <- species_accumulation(subspecies_data, "subspecies")
    boxplot_data_g <- species_accumulation(genera_data, "genera")

  
    accum_s_long <- pivot_longer(boxplot_data_s, cols=everything(),
                                 names_to = "Samples", 
                                 values_to = "Richness_value") %>%
      mutate(SampleNum = as.numeric(gsub("V", "", Samples)), Taxa = "species")
  
    accum_ss_long <- pivot_longer(boxplot_data_ss, cols=everything(), 
                                  names_to = "Samples", values_to = "Richness_value") %>%
      mutate(SampleNum = as.numeric(gsub("V", "", Samples)), Taxa= "subspecies")
    
    accum_g_long <- pivot_longer(boxplot_data_g, cols=everything(),
                                 names_to = "Samples", values_to = "Richness_value") %>%
      mutate(SampleNum = as.numeric(gsub("V", "", Samples)), Taxa= "genera")
    
    accum_merge <- rbind(as.data.frame(accum_s_long), as.data.frame(accum_ss_long),
                         as.data.frame(accum_g_long))

    accum_plot <- ggplot(data=accum_merge, aes(x=factor(SampleNum), y=Richness_value, fill=Taxa))+
      scale_fill_manual(values = other_colors) +
      guides(fill= guide_legend(position="bottom")) +
      geom_boxplot(size=0.1, width=1) +
      labs(x="Number of samples evaluated", y="Cumulative number of taxa detected") +
      scale_x_discrete(breaks=(c(1,10,20,30,40,50,60, 70, 80, 90, 95))) +
      theme_bw() +
      theme(text=element_text(size=12))

  return(accum_plot)
}

genus_df_pas <- RPKMF_rank(esviritu_passive_filt,"genus")
species_df_pas <- RPKMF_rank(esviritu_passive_filt,"species")
subspecies_df_pas <- RPKMF_rank(esviritu_passive_filt, "subspecies")

Fig_S1D <- plot_S1D(species_df_pas,
                               subspecies_df_pas, 
                               genus_df_pas)
Fig_S1D

ggsave(Fig_S1D, 
       file = file.path(data_dir, "Fig_S1D.jpg"), 
       width = 6, height = 4, units = "in", dpi=300)
```


### Tax tree


```{r}
tax_hierarchy_hc <- esviritu_passive_filt %>%
  distinct(species, genus, family,
           order, tclass, phylum) %>%
  ungroup() %>%
  mutate(species = gsub("\\W+", "", species)) %>%
  mutate(genus = gsub("\\W+", "", genus)) %>%
  mutate(family = gsub("\\W+", "", family)) %>%
  mutate(order = gsub("\\W+", "", order)) %>%
  mutate(tclass = gsub("\\W+", "", tclass)) %>%
  mutate(phylum = gsub("\\W+", "", phylum))
``` 

```{r}
tax_hierarchy_hc_df <- as.data.frame(tax_hierarchy_hc)


tax_hierarchy_hc_df <- as.data.frame(unclass(tax_hierarchy_hc_df),
                                 stringsAsFactors=TRUE,
                                 row.names = "species")

cols <- colnames(tax_hierarchy_hc_df)

taxo_phylo_hc <- taxo2phylog(as.taxo(tax_hierarchy_hc_df))
```

```{r}
species_meta_dt <- esviritu_passive_filt %>%
  group_by(Site_cat) %>%
  #count total_samples as the number of unique sample names
  mutate(total_samples = n_distinct(sample_ID)) %>%
  ungroup() %>%
  #count how many distinct samples contained each species (n)
  group_by(species, Site_cat, total_samples) %>%
  summarize(n = n_distinct(sample_ID)) %>%
  mutate(freq = n / total_samples) %>%
  ungroup() %>%
  complete(nesting(Site_cat, total_samples), species,
           fill=list(n=0, freq=0))

freq_info_dt <- species_meta_dt %>%
  group_by(species) %>%
  mutate(total_number = sum(n),
         h_freq = freq/ sum(freq)) %>%
  select(species, total_number, h_freq) %>%
  ungroup() %>%
  distinct() %>%
  merge(., esviritu_passive_filt[, c("species", "family")], by="species", all.x = TRUE,
        all.y = FALSE) %>%
  mutate(species = gsub("\\W+", "", species))
```

```{r warning=FALSE}
tree <- ggtree::ggtree(taxo_phylo_hc, layout = "daylight",branch.length = "scaled",
             color="grey40", size=0.2)
```

Get the annotation commands for easier plotting:

```{r}
get_clade_annot_commands <- function(tax.phy_obj, search_data, tax_tree_data){
  #get the family nodes
  third_nodes <- sapply(tax.phy_obj$paths, function(path) {
  if (length(path) >= 5) path[5] else NA
  })

  #use only the nodes with s__ on the name
  third_nodes_idx <- grep("s__", names(third_nodes))
  #third_nodes <- na.omit(third_nodes)
  #return(((third_nodes[third_nodes_idx])))
  clade_commands <- sapply(seq_along(third_nodes[third_nodes_idx]), function(i) {
    f_node_num <- tax_tree_data@data[tax_tree_data@data$label == unname(third_nodes[i]), "node"]
    node_label <- unique(search_data[search_data$species == names(third_nodes[i]), "family"])
    paste0("geom_cladelab(node = ", f_node_num, 
           ", label = '", node_label, 
           "', fontsize = 5) +")
  })

  clade_commands <- unique(clade_commands)
  all_clades <- paste(clade_commands, collapse="\n")
  return(all_clades)
}
```


```{r}
#create a search table from esviritu results
species_family_df <- esviritu_passive_filt[, c("family", "species")] %>%
  mutate(family=gsub("f__", "", family),
         species = gsub("\\W+", "", species))

annots <- get_clade_annot_commands(taxo_phylo_hc, species_family_df, tree)
#str(annots)
(cat(annots))

```


```{r}
Fig_S1E <- tree %<+% freq_info_dt + 

    geom_tippoint(aes(color=family, size = total_number), alpha = 0.7) +
  scale_size_continuous(name = "Total Detections", breaks = c(1, 10, 25, 50, 95)) +
  scale_color_manual(values=colorRampPalette(other_colors[1:9])(33)) +

  guides(size = guide_legend("Total Detections (1-95)", order=1),
         color="none") +
  
  geom_cladelab(node = 562, label = 'Hepadnaviridae', fontsize = 5, hjust=0) +
  geom_cladelab(node = 566, label = 'Retroviridae', fontsize = 5, hjust=-0.05) +
  geom_cladelab(node = 574, label = 'Polyomaviridae', fontsize = 5, hjust=-0.05) +
  geom_cladelab(node = 581, label = 'Papillomaviridae', fontsize = 5, hjust=-0.1) +
  geom_cladelab(node = 599, label = 'Parvoviridae', fontsize = 5, hjust=-0.1, vjust=-0.5) +
  geom_cladelab(node = 621, label = 'Circoviridae', fontsize = 5, hjust=0.5, vjust=-0.7) +
  geom_cladelab(node = 626, label = 'Smacoviridae', fontsize = 5, hjust=1 ,vjust=0) +
  geom_cladelab(node = 630, label = 'Genomoviridae', fontsize = 5, hjust=1.1, vjust=1) +
  geom_cladelab(node = 638, label = 'Sedoreoviridae', fontsize = 5, hjust=1.1) +
  geom_cladelab(node = 641, label = 'Spinareoviridae', fontsize = 5, hjust=1.05) +
  geom_cladelab(node = 647, label = 'Hepeviridae', fontsize = 5, hjust=1.1, vjust=1, barsize=0) +
  
  geom_taxalink(taxa1="s__Paslahepevirusbalayani",taxa2="s__HepatitisEvirus",
                color="black", curvature=0.4, offset=-0.01, ncp=3) +
  
  geom_cladelab(node = 652, label = 'Virgaviridae', fontsize = 5, hjust=1.1) +
  geom_cladelab(node = 656, label = 'Flaviviridae', fontsize = 5, hjust=1) +
  geom_cladelab(node = 661, label = 'Tombusviridae', fontsize = 5, hjust=1.05) +
  geom_cladelab(node = 672, label = 'Phenuiviridae', fontsize = 5, hjust=1) +
  geom_cladelab(node = 676, label = 'Orthomyxoviridae', fontsize = 5, hjust=2.2, vjust=-1, barsize=0) +
  
  geom_taxalink(taxa1="s__Cytomegalovirushumanbeta5",taxa2="s__Varicellovirushumanalpha3",
                color="black", curvature=0.4, offset=-0.02, ncp=3) +
  
  geom_cladelab(node = 682, label = 'Paramyxoviridae', fontsize = 5, hjust=1) +
  geom_cladelab(node = 685, label = 'Pneumoviridae', fontsize = 5, hjust=1) +
  geom_cladelab(node = 691, label = 'Poxviridae', fontsize = 5, hjust=1) +
  geom_cladelab(node = 696, label = 'Orthoherpesviridae', fontsize = 5, hjust=2.05, vjust=3, barsize=0) +
  
  geom_taxalink(taxa1="s__Alphainfluenzavirusinfluenzae",taxa2="s__Gammainfluenzavirusinfluenzae",
                color="black", curvature=0.4, offset=-0.02, ncp=3) +
  
  geom_cladelab(node = 708, label = 'Picobirnaviridae', fontsize = 5, hjust=1) +
  geom_cladelab(node = 714, label = 'Coronaviridae', fontsize = 5, hjust=1.1, vjust=0.1) +
  geom_cladelab(node = 720, label = 'Mesoniviridae', fontsize = 5, hjust=1.1, vjust=-0.1) +
  geom_cladelab(node = 722, label = 'Tobaniviridae', fontsize = 5, hjust=1.1, vjust=0.2) +
  geom_cladelab(node = 725, label = 'Caliciviridae', fontsize = 5, hjust=1, vjust=1) +
  geom_cladelab(node = 728, label = 'Dicistroviridae', fontsize = 5, hjust=1, vjust=1) +
  geom_cladelab(node = 733, label = 'Iflaviridae', fontsize = 5,hjust=1, vjust= 1.3) +
  geom_cladelab(node = 736, label = 'Picornaviridae', fontsize = 5, vjust=1.2, hjust=0.2) +
  geom_cladelab(node = 746, label = 'Secoviridae', fontsize = 5, vjust=1) +
  geom_cladelab(node = 754, label = 'Solemoviridae', fontsize = 5, hjust=-0.1) +
  geom_cladelab(node = 759, label = 'Astroviridae', fontsize = 5, hjust=-0.1) +
  geom_cladelab(node = 766, label = 'Adenoviridae', fontsize = 5, hjust=-0.05) +
  geom_cladelab(node = 774, label = 'Anelloviridae', fontsize = 5, hjust=-0.05) +
  
  xlim(-15,15) +

  theme(legend.position = "bottom", 
        legend.direction = "horizontal",
        #legend.box = "vertical",
        legend.title = element_text(size = 10),
        legend.text = element_text(size=10),
        panel.background = element_rect(fill = "transparent", color = NA),
        plot.background = element_rect(fill = "transparent", color = NA),
        legend.background = element_rect(fill = "transparent", color = NA),
        legend.box.background = element_rect(fill = "transparent", color = NA)) 

Fig_S1E

ggsave(Fig_S1E, file= file.path(data_dir, "Fig_S1E.jpg"), 
                                  width=15, height=6, dpi=300)
```

## Figure 1

### A, B and C viral read plots

```{r}
#plot viral read proportion of all samples
plot_Fig1A <- function(data, colors){
  ggplot(data=data, aes(x=factor(Site_cat, levels=c('UH', 'WWTP', 'SA')), 
                        y=viral_reads_percentage, fill=Site_cat)) +
    geom_boxplot(staplewidth = 0.5,
                 outliers = FALSE) +
    geom_jitter(width=0.25, alpha=0.8) +
    scale_fill_manual(values=colors) +
    theme_bw()+
    labs(y = "Percentage",
         x="") +
    theme(axis.text.x = element_text(size=14),
          axis.text.y = element_text(size=14),
          axis.title.y = element_text(size=14),
          legend.position = "none",
          panel.grid = element_line(color="white")) + 
    scale_y_continuous(labels = function(x) paste0(x, "%"))

}


#plot number of viral reads for all samples
plot_Fig1B <- function(data, colors){
  ggplot(data=data, aes(x=factor(Site_cat, levels=c('UH', 'WWTP', 'SA')), 
                        y=total_mapped, fill=Site_cat)) +
    geom_boxplot(staplewidth = 0.5,
                 outliers = FALSE) +
    geom_jitter(width=0.25,alpha=0.8) +
    scale_fill_manual(values=colors) +
    theme_bw()+
    labs(y = "Reads aligned to viral genomes (N)",
         x="") +
    scale_y_log10() +
    theme(axis.text.x = element_text(size=14),
          axis.title.y = element_text(size=14),
          axis.text.y = element_text(size=14),
          legend.position = "none",
          panel.grid = element_line(color="white"))
}


#plot number of reads (total filtered) for all samples
plot_Fig1C <- function(data, colors){

  ggplot(data=data, aes(x=factor(Site_cat, levels=c('UH', 'WWTP', 'SA')), 
                        y=total_reads, fill=Site_cat)) +
    geom_boxplot(staplewidth = 0.5,
                 outliers = FALSE) +
    geom_jitter(width=0.25,alpha=0.8) +
    scale_fill_manual(values=colors) +
    theme_bw()+
    labs(y = "Total number of filtered reads in sample",
         x="") +
    scale_y_log10() +
    theme(axis.text.x = element_text(size=14),
          axis.text.y = element_text(size=14),
          axis.title.y = element_text(size=14),
          legend.position = "none",
          panel.grid = element_line(color="white"))
}

```


```{r}
v_reads <- summarize_viral_reads(esviritu_passive_filt)

v_reads_meta <- merge(v_reads, esviritu_passive_filt[, c("sample_ID", "Site_cat")] %>% distinct(), by="sample_ID")

Fig1A <- plot_Fig1A(v_reads_meta, site_cat_colors) 

Fig1B <- plot_Fig1B(v_reads_meta, site_cat_colors) 

Fig1C <- plot_Fig1C(v_reads_meta, site_cat_colors)


plot_grid(Fig1A, Fig1B, Fig1C, ncol=3)


ggsave(plot_grid(Fig1A, Fig1B, Fig1C, ncol=3), 
       file = file.path(data_dir, "Fig1_ABC.jpg"), 
       width=15, height=5, units = "in", dpi=300)

```
Kruskal-Wallis tests and Dunn's post hoc test with Bonferroni corrections
```{r}
kruskal.test(v_reads_meta$total_reads, v_reads_meta$Site_cat)
kruskal.test(v_reads_meta$total_mapped, v_reads_meta$Site_cat)
kruskal.test(v_reads_meta$viral_reads_percentage, v_reads_meta$Site_cat)
```
```{r}
DunnTest(v_reads_meta$total_mapped, v_reads_meta$Site_cat,
         method=c("bonferroni"), alternative="two.sided")

DunnTest(v_reads_meta$viral_reads_percentage, v_reads_meta$Site_cat,
         method=c("bonferroni"), alternative="two.sided")
```

### D Rarefaction plot

```{r}
read_matrix <- function(data, taxa_level){
  rm <- data %>%
    group_by(sample_ID, !!sym(taxa_level)) %>%
    summarize(reads=sum(read_count)) %>%
    ungroup()

  rm <- as.data.frame(pivot_wider(data=rm,
                    names_from = sample_ID,
                    values_from = reads,
                    values_fill = list(reads=0)))

  return(rm)

}

plot_Fig1D <- function(data, taxa_level){
  read_counts <- read_matrix(data, taxa_level)
  
  rarecurve_data <- read_counts[,-1]
  rownames(rarecurve_data) <- read_counts[[taxa_level]]
 
  rarecurve <-  rarecurve(t(rarecurve_data), step = 500, 
                          label=FALSE, cex=0.2, tidy = TRUE)
  
  colnames(rarecurve) <- c("sample_ID", "Sample", "Taxa_level")
  
  plot_data <-  merge(rarecurve, esviritu_passive_filt[, c("sample_ID", "Site_cat")] %>% distinct(), by="sample_ID")

  plot <- ggplot(data=plot_data, aes(x=Sample, y=Taxa_level, color=Site_cat)) +
    geom_point(size=0.1,  alpha=0.7) +
    theme_bw() +
    labs(x="Number of sequences",
        y="Number of species") +
    scale_color_manual(values = site_cat_colors , name="") +
    theme(legend.position = "right") +
    guides(color = guide_legend(override.aes = list(size = 5),
                                position="bottom"))
  
  return(plot)
}
```

```{r}
Fig1D <- plot_Fig1D(esviritu_passive_filt, "species")
Fig1D

ggsave(Fig1D, file = file.path(data_dir, "Fig1_D.jpg"), 
       width = 6, height = 4, dpi=300)
```

### E, F and G, Species count, Shannon and Simpson

```{r functions-for-adiv-calculation}
summarize_tax_levels <- function(data){
  data %>%
    group_by(sample_ID) %>%
    summarize(phylums=n_distinct(phylum),
              classes=n_distinct(tclass),
              orders=n_distinct(order),
              families=n_distinct(family),
              genera=n_distinct(genus),
              species=n_distinct(species),
              subspecies=n_distinct(subspecies),
              .groups="drop")
}

RPKMF_df <- function(data, tax_level, RPKMF_type){
  RPKMF_df <- data %>%
    group_by(sample_ID, !!sym(tax_level)) %>%
    summarize(RPKMF = sum(!!sym(RPKMF_type))) %>%
    ungroup()
  
  RPKMF_df_wide <-pivot_wider(data=RPKMF_df,
                    names_from = sample_ID,
                    values_from = RPKMF,
                    values_fill = list(RPKMF=0))

  return(RPKMF_df_wide)
}
```

```{r alpha-diversity-calculation}
#calculate alpha diversity metrics for species level data
alpha_diveristy_calc <- function(data, RPKMF_type){
  
  species_df <- RPKMF_df(data, "species", RPKMF_type)
  adiv <- data.frame(sample_ID=colnames(species_df[,-1]),
                  shannon_index = diversity(x=t(species_df[,-1]), 
                                            index="shannon",
                                            equalize.groups = FALSE),
                  simpson_index_inv = diversity(x=t(species_df[,-1]), 
                                            index="invsimpson",
                                            equalize.groups = FALSE),
                  species_count = summarize_tax_levels(data)$species)

  #add metadata to alpha diversity 
  adiv <- merge(adiv, esviritu_passive_filt[, c("sample_ID", "Site_cat", "filtered_reads_in_sample")] %>% distinct(), by="sample_ID")

  return(adiv)
  
}

alpha_div_passive <- alpha_diveristy_calc(esviritu_passive_filt, "RPKMF_norm")
```

```{r plots-of-alphadiv-distributions}
Fig1E <- ggplot(data = alpha_div_passive, 
                  aes(x=factor(Site_cat, levels=c('UH', 'WWTP', 'SA')), 
                      y=species_count, fill=Site_cat)) +
  geom_boxplot(staplewidth = 0.5,
                 outliers = FALSE) +
  geom_jitter(width=0.25,alpha=0.8) +
  scale_fill_manual(values=site_cat_colors) +
  guides(fill="none") +
  labs(x="", y="Species count (N)") +
  theme_bw() +
    theme(axis.text.x = element_text(size=14),
          axis.text.y = element_text(size=14),
          axis.title.y = element_text(size=14),
          legend.position = "none",
          panel.grid = element_line(color="white"))

Fig1F <- ggplot(data = alpha_div_passive, 
                  aes(x=factor(Site_cat, levels=c('UH', 'WWTP', 'SA')), 
                      y=shannon_index, fill=Site_cat)) +
  geom_boxplot(staplewidth = 0.5,
                 outliers = FALSE) +
  geom_jitter(width=0.25,alpha=0.8) +
  scale_fill_manual(values=site_cat_colors)+
  guides(fill="none") +
  labs(x="", y="Shannon index") +
  theme_bw()+
    theme(axis.text.x = element_text(size=14),
          axis.text.y = element_text(size=14),
          axis.title.y = element_text(size=14),
          legend.position = "none",
          panel.grid = element_line(color="white"))

Fig1G <- ggplot(data = alpha_div_passive, 
                  aes(x=factor(Site_cat, levels=c('UH', 'WWTP', 'SA')), 
                      y=simpson_index_inv, fill=Site_cat)) +
  geom_boxplot(staplewidth = 0.5,
                 outliers = FALSE) +
  geom_jitter(width=0.25,alpha=0.8) +
  scale_fill_manual(values=site_cat_colors)+
  guides(fill="none") +
  labs(x="", y="Simpson index (1/D)") +
  theme_bw()+
    theme(axis.text.x = element_text(size=14),
          axis.text.y = element_text(size=14),
          axis.title.y = element_text(size=14),
          legend.position = "none",
          panel.grid = element_line(color="white"))

Fig1E
Fig1F
Fig1G

ggsave(plot_grid(Fig1E, Fig1F, Fig1G, ncol=3), 
       file = file.path(data_dir, "Fig1_EFG.jpg"), 
       width=15, height=5, units = "in", dpi=300)
```

Kruskal-Wallis and Dunn's post hoc test for alpha diversity metrics

```{r}
kruskal.test(alpha_div_passive$species_count, alpha_div_passive$Site_cat)
kruskal.test(alpha_div_passive$shannon_index, alpha_div_passive$Site_cat)
kruskal.test(alpha_div_passive$simpson_index_inv, alpha_div_passive$Site_cat)
```


```{r}
DunnTest(alpha_div_passive$species_count, alpha_div_passive$Site_cat,
         method=c("bonferroni"), alternative="two.sided")

DunnTest(alpha_div_passive$shannon_index, alpha_div_passive$Site_cat,
         method=c("bonferroni"), alternative="two.sided")

DunnTest(alpha_div_passive$simpson_index_inv, alpha_div_passive$Site_cat,
         method=c("bonferroni"), alternative="two.sided")
```

### H Bray-Curtis PCoA

```{r PCoA-Bray-Curtis}
plot_Fig1H <- function(data, taxa_level, RPKMF_type){

  #calculate RPKMF df (cols species, rows samples)
  RPKMF_df <- RPKMF_df(data, taxa_level, RPKMF_type)
  
  bray_curtis <- as.data.frame(as.matrix(vegdist(t(RPKMF_df[,-1]))), method="bray")
  
  bray_curtis$sample_ID1 <- rownames(bray_curtis)

  pcoa <- cmdscale(bray_curtis[,-ncol(bray_curtis)], eig=TRUE)
  pcoa_df <- as.data.frame(pcoa$points)
  pcoa_df$sample_ID <- bray_curtis$sample_ID1

  pcoa_df_meta <- merge(pcoa_df, esviritu_passive_filt[, c("sample_ID", "Site_cat", "City")] %>% distinct(), by="sample_ID")
  
  #explained variance by PC 
  
  pcoa_eig <- pcoa$eig
  pcoa_eig_pos <- pcoa_eig[pcoa_eig>0]
  
  exp_var <- pcoa_eig_pos /(sum(pcoa_eig_pos))
  #return(exp_var)
  
  PC1_var <- round(exp_var[1]*100, 2)
  PC2_var <- round(exp_var[2]*100, 2)

  plot <- ggplot() +
  geom_vline(xintercept=0, color="grey80", alpha=0.8) +
  geom_hline(yintercept=0, color="grey80", alpha=0.8) +
  
  geom_point(data=pcoa_df_meta, 
             aes(x = V1, y = V2, 
                 color=Site_cat,
                 shape=City, group=Site_cat), 
             size=4, alpha=0.9) +
    
  labs(x=sprintf("Bray-Curtis PCoA1 (%s %%)", PC1_var), 
       y= sprintf("Bray-Curtis PCoA2 (%s %%)", PC2_var)) +
  scale_color_manual(values=site_cat_colors, name="Location") + 
  scale_shape_manual(values=c(15, 16), name="City") +
  guides(color=guide_legend(position="bottom"),
         shape=guide_legend(position="bottom")) +
  theme_bw()+
    theme(panel.grid = element_line(color="white")) 

}

```

```{r}
Fig1H <- plot_Fig1H(data=esviritu_passive_filt, 
                       taxa_level="species", RPKMF_type = "RPKMF_norm")


Fig1H

ggsave(Fig1H, file=file.path(data_dir, "Fig1_H.jpg"), 
                                  width=6, height=4, dpi=300)
```

PERMANOVA on Bray-Curtis distance
```{r}
RPKMF_data <- RPKMF_df(esviritu_passive_filt, "species", "RPKMF_norm")
bray_curtis <- as.data.frame(as.matrix(vegdist(t(RPKMF_data[,-1]))), method="bray")

meta_bc <- esviritu_passive_filt %>%
  select(sample_ID, Site_cat) %>%
  distinct() %>%
  as.data.frame()

rownames(meta_bc) <- meta_bc$sample_ID

meta_bc <- meta_bc[rownames(bray_curtis), ]

permanova_bc <- adonis2(bray_curtis ~ Site_cat, data=meta_bc,
                                    permutations=999, by="margin")

permanova_bc
```


```{r}
bd <- betadisper(vegdist(t(RPKMF_data[,-1]), method="bray"), meta_bc$Site_cat)

permutest(bd, permutations = 999)
```



### I stacked bar plot of family level detections

```{r}
set.seed(26) #set seed to get right colors

barplot_data <- esviritu_passive_filt %>%
 # merge(., esviritu_passive_filt[, c("sample_ID", "Site_cat")] %>% distinct(), by="sample_ID") %>%
  mutate(family =gsub("f__", "", family)) %>%
  filter(!is.na(family)) %>%
  group_by(Site_cat, family) %>%
  summarize(RPKMF_norm = sum(RPKMF_norm), .groups="drop") %>%
  ungroup() %>%
  group_by(Site_cat) %>%
  arrange(desc(RPKMF_norm), .by_group = TRUE) %>%
    mutate(family = ifelse(row_number() > 8, "Others", family)) %>%
    group_by(Site_cat, family) %>%
    summarise(total_value = sum(RPKMF_norm), .groups = "drop")


family_order <- barplot_data %>%
  group_by(family) %>%
  summarize(RPKMF_sum=sum(total_value)) %>%
  arrange(desc((RPKMF_sum))) %>%
  pull(family)

family_ordered_top <- c(family_order[family_order != "Others"], "Others")
  
barplot_data$family <- factor(barplot_data$family,
                              levels = rev(family_ordered_top))


colors <- sample(colorRampPalette(other_colors[1:9])(10))
names(colors) <- unique(barplot_data %>% filter(family != "Others" & !is.na(family)) %>% pull(family))

colors_plus_other <- c(colors, "Others" = "grey60")


Fig1I <- ggplot(barplot_data, aes(fill=family, y=total_value, x=factor(Site_cat, levels=c("UH", "WWTP", "SA")))) + 
    geom_bar(position="fill", stat="identity") +
  scale_fill_manual(values=colors_plus_other) +
  guides(fill = guide_legend(reverse = TRUE,
                             position="bottom", ncol=5)) +
  theme_bw() +
  theme(legend.spacing.y = unit(0, "cm"),
        legend.key.height = unit(0.2, "cm"),
        legend.text = element_text(size=14),
        axis.text.x = element_text(size=14),
        axis.text.y = element_text(size=14),
        legend.title = element_text(size=14)) +
  labs(x="", y="Relative normalized RPKMF") +
  scale_y_continuous(labels = function(x) paste0(x*100, "%"))

Fig1I

ggsave(Fig1I, file=file.path(data_dir, "Fig1_I.jpg"), 
                                  width=9, height=8, dpi=300)
```
### J and K Host distribution by accession and RPKMF

```{r functions-for-pie-charts}
summarize_tax_levels <- function(data, col, taxa_level){
  data <- data %>%
    group_by(!!sym(col)) %>%
    summarize(phylums=n_distinct(phylum),
              classes=n_distinct(tclass),
              orders=n_distinct(order),
              families=n_distinct(family),
              genera=n_distinct(genus),
              species=n_distinct(species),
              subspecies=n_distinct(subspecies),
              Accession=n_distinct(Accession),
              .groups="drop") %>%
    mutate(total_value=!!sym(taxa_level)) %>%
    rename(category=!!sym(col)) %>%
    filter(category != "" & !is.na(category)) %>%
    arrange(desc(total_value)) %>%
    mutate(category = ifelse(row_number() > 7 | category=="" , "Others", category)) %>%
    group_by(category) %>%
    summarize(total_value=sum(total_value), .groups="drop") %>%
    mutate(proportion = 100 * total_value / sum(total_value)) %>%
    arrange((total_value))
  
  return(data)
  
}

sum_cat_top_N <- function(data, col, val){
  data <- data %>%
    group_by(!!sym(col)) %>%
    summarize(total_value=sum(!!sym(val)), .groups="drop") %>%
    rename(category = !!sym(col)) %>%
    filter(category != "" & !is.na(category)) %>%
    arrange(desc(total_value)) %>%
    mutate(category = ifelse(row_number() > 7 | category=="", "Others", category)) %>%
    group_by(category) %>%
    summarise(total_value = sum(total_value), .groups = "drop")%>% 
    mutate(proportion = 100 * total_value / sum(total_value),
         category=category) %>%
    arrange((total_value))
  
  return(data)
  
}
```

```{r}
host_acc_UH <- summarize_tax_levels(esviritu_passive_filt %>% filter(Site_cat=="UH"), 
                                    "host_genus", "Accession") %>%
  mutate(Site_cat="UH",
         datatype="Accesion") 

host_rpkmf_UH <- sum_cat_top_N(esviritu_passive_filt %>% filter(Site_cat=="UH"), 
                               "host_genus", "RPKMF_norm") %>%
  mutate(Site_cat="UH",
         datatype="RPKMF") 

host_acc_SA <- summarize_tax_levels(esviritu_passive_filt %>% filter(Site_cat=="SA"), 
                                    "host_genus", "Accession") %>%
  mutate(Site_cat="SA",
         datatype="Accesion") 

host_rpkmf_SA <- sum_cat_top_N(esviritu_passive_filt %>% filter(Site_cat=="SA"), 
                               "host_genus", "RPKMF_norm") %>%
  mutate(Site_cat="SA",
         datatype="RPKMF") 

host_acc_WWTP <- summarize_tax_levels(esviritu_passive_filt %>% filter(Site_cat=="WWTP"), 
                                      "host_genus", "Accession") %>%
  mutate(Site_cat="WWTP",
         datatype="Accesion") 

host_rpkmf_WWTP <- sum_cat_top_N(esviritu_passive_filt %>% filter(Site_cat=="WWTP"), 
                                 "host_genus", "RPKMF_norm") %>%
  mutate(Site_cat="WWTP",
         datatype="RPKMF")
```

```{r}
host_colors <- c("Homo" = "#8A6399",
                 "Others" = "grey60",
                 "Sus" = "#197C3F",
                 "Gallus" = "#3B459F",
                 "Bos" = "#7BC3DA",
                 "Canis" = "#DFD43D",
                 "Felis" = "#D45F0B",
                 "Capsicum" = "#A47E58",
                 "Equus" = "#4E95D4",
                 "Culex" = "#948D28",
                 "Vitis" = "#456DBA",
                 "Solanum" = "#a1c999",
                 "Contamination" = "#CC79A7",
                 "Cucurbita" = "#D1683F",
                 "Hipposideros" = "#A64799",
                 "Aphis" = "#D6DE6B")
```


```{r}
pie_data_ACC <- rbind(host_acc_UH, 
                  host_acc_SA, 
                  host_acc_WWTP) %>%
  group_by(Site_cat) %>%
  arrange(desc(total_value), .by_group = TRUE) 

pie_data_ACC$category <- forcats::fct_reorder(
  pie_data_ACC$category,
  pie_data_ACC$total_value,
  .fun = sum,
  .desc = FALSE
)

pie_data_ACC$category <- pie_data_ACC$category |>
  forcats::fct_reorder(
    pie_data_ACC$total_value,
    .fun = sum,
    .desc = FALSE
  )

pie_data_ACC$category <- forcats::fct_relevel(pie_data_ACC$category, "Others")

Fig_1J <- ggplot(data = pie_data_ACC,
       aes(y = factor(Site_cat, levels=c("SA", "WWTP", "UH")), x = total_value, fill = category)) +
  geom_bar(position="fill", stat="identity", width = 0.7) +
  scale_fill_manual(values=host_colors) +
  #scale_fill_manual(values=sample(colorRampPalette(other_colors[1:9])(20))) +
  scale_x_continuous(labels = function(x) paste0(x*100, "%")) +
   theme_bw() +
  theme(panel.spacing.y = unit(0, "lines")) +
  guides(fill = guide_legend(title="Host Genus", reverse = TRUE)) +
  labs(x="Relative Accession counts", y="")

Fig_1J

ggsave(fig_1J, file= file.path(data_dir, "Fig1_J.jpg"), 
                                  width=10, height=4, dpi=300)
```

```{r}
pie_data_RPKMF <- rbind(host_rpkmf_UH,
                  host_rpkmf_SA,
                  host_rpkmf_WWTP) %>%
  group_by(Site_cat) %>%
  arrange(desc(total_value), .by_group = TRUE) 

pie_data_RPKMF$category <- pie_data_RPKMF$category |>
  forcats::fct_reorder(
    pie_data_RPKMF$total_value,
    .fun = sum,
    .desc = FALSE
  )

pie_data_RPKMF$category <- forcats::fct_relevel(pie_data_RPKMF$category, "Others")



Fig_1K <- ggplot(data = pie_data_RPKMF,
       aes(y = factor(Site_cat, levels=c("SA", "WWTP", "UH")), x = total_value, fill = category)) +
  geom_bar(position="fill", stat="identity", width = 0.7) +
  scale_fill_manual(values=host_colors) +
  scale_x_continuous(labels = function(x) paste0(x*100, "%")) +
   theme_bw() +
  theme(panel.spacing.y = unit(0, "lines")) +
  guides(fill = guide_legend(title="Host Genus", reverse = TRUE)) +
  labs(x="Relative normalized RPKMF", y="")

Fig_1K

ggsave(fig_1K, file= file.path(data_dir, "Fig1_K.jpg"), 
                                  width=10, height=4, dpi=300)
```
